Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2020-09-22 Manuscript/HI_Model_ms.Rmd
updated manuscript files

### *CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@Walker_integrating_2020]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@Walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model ecosystem carbon balance projections to the influence of differing rates of a CO~2~ fertilization effect (CFE). We incorporated a NPP CFE multiplier representing the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

### *Scenario simulations and analysis*
Each of the four unique scenarios were run for 90 years at an annual timestep and repeated for 30 Monte Carlo realizations, using initial conditions corresponding to the year 2010. All simulations were performed within the SyncroSim (version 2.2.4) software framework with ST-Sim (version 3.2.13) and SF (version 3.2.10) add-on modules [@daniel_state-and-transition_2016; @daniel_integrating_2018]. Model inputs and outputs were prepared with the R statistical computing platform [@RCore_2019] using the tidyverse [@Wickham_tidyverse_2019], raster [@hijmans_raster_2020], and rsyncrosim [@daniel_rsyncrosim_2020] packages. Carbon stocks and fluxes for the seven main Hawaiian Islands were calculated for each scenario by summing within each of the 30 Monte Carlo realizations on an annual basis and then calculating annual means as well as the annual upper and lower limits of the 30 Monte Carlo realizations. Carbon balance for the seven main Hawaiian Islands was calculated on annual basis for each scenario and Monte Carlo realization as net biome productivity (NBP), which was equal to annual carbon input in the form of net primary productivity (NPP) minus the annual sum of all carbon losses from terrestrial ecosystems, including heterotrophic respiration (R~h~) from litter and soil, carbon fluxes to the atmosphere triggered by land use and land-use change, wildfire emissions, and aquatic carbon losses through leaching and overland flow. Positive NBP values indicated ecosystems of the seven main Hawaiian Islands were acting as a net sink for atmospheric CO~2~, while negative NBP values indicated that these ecosystems were acting as a net carbon source to the atmosphere [@chapin_reconciling_2006].

@@

Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land-use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (figure S4) and sustained losses of forest and shrubland land area in the high land-use scenario (figure S5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (figure S4) resulted in R~h~ being ~3% higher by 2100 than under RCP 4.5 (figure 3).  Land-use change substantially reduced R~h~ in the high land-use scenario (figure 3) because of long-term reductions in forest and shrubland land area (figure S5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land-use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land-use scenario and an average of ~0.2 Tg y^-1^ in the low land-use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land-use scenario, driven by greater variability in wildland fire probabilities.

@@

### *Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined under all four scenarios over time (figure 4).  On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). In addition, the high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reduction and avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

@@

### *CO~2~ fertilization effect*
Projected estimates of both total ecosystem carbon storage and ecosystem carbon balance were highly sensitive to differing rates of a CFE on plant productivity. Under the high radiative forcing (RCP 8.5) and high land-use scenario, the inclusion of a CFE ranging from 5-15% led to ~33-98 Tg of additional carbon storage in ecosystems by the end of the century, a ~10-30% increase (figure 6a). Compared to the reference scenario (0% CFE), a 5% CFE was sufficient to transform Hawaiian Island ecosystems from a net carbon source to the atmosphere for the latter half of the 21st century (figure 4b) to a net carbon sink for the entire simulation period (figure 6b), completely offsetting all other carbon losses induced by high radiative forcing and high land use. Net carbon sink strength was further enhanced at higher CFE rates, with NBP increasing by an average of 0.07 Tg C y^-1^ for each 1% increase in CFE (figure 6b). When compared to other scenarios, the high radiative forcing and high land-use scenario with a 5% CFE resulted in a mean annual NBP of 0.46 ± 0.3 Tg C y^-1^,  roughly equivalent to mean annual NBP of the low radiative forcing and low land-use scenario with no CFE (0.52 ± 0.12). A 15% CFE applied to the high radiative forcing and high land-use scenario resulted in a mean annual NBP  of 1.18 ± 0.29 Tg C y^-1^, more than double that of the low radiative forcing and low land-use scenario with no CFE.

>>>>

### *CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@Walker_integrating_2020]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@Walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model ecosystem carbon balance projections to  different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

### *Scenario simulations and analysis*
Each of the four unique scenarios were run for 90 years at an annual timestep and repeated for 30 Monte Carlo realizations, using initial conditions corresponding to the year 2010. All simulations were performed within the SyncroSim (version 2.2.4) software framework with ST-Sim (version 3.2.13) and SF (version 3.2.10) add-on modules [@daniel_state-and-transition_2016; @daniel_integrating_2018]. Model inputs and outputs were prepared with the R statistical computing platform [@RCore_2019] using the tidyverse [@Wickham_tidyverse_2019], raster [@hijmans_raster_2020], and rsyncrosim [@daniel_rsyncrosim_2020] packages. Carbon stocks and fluxes for the seven main Hawaiian Islands were calculated for each scenario by summing within each Monte Carlo realization on an annual basis and then calculating annual means as well as the annual upper and lower limits of the 30 Monte Carlo realizations. Carbon balance for the seven main Hawaiian Islands was calculated on annual basis for each scenario and Monte Carlo realization as net biome productivity (NBP), which was equal to annual carbon input in the form of NPP minus the annual sum of all carbon losses from terrestrial ecosystems, including heterotrophic respiration (R~h~) from litter and soil, carbon fluxes to the atmosphere triggered by land use and land-use change, wildfire emissions, and aquatic carbon losses through leaching and overland flow. Positive NBP values indicated ecosystems of the seven main Hawaiian Islands were acting as a net sink for atmospheric CO~2~, while negative NBP values indicated that these ecosystems were acting as a net carbon source to the atmosphere [@chapin_reconciling_2006].

@@

Net primary production (NPP) for the seven main Hawaiian Islands declined across all four scenarios, driven primarily by climate change and to a lesser extent by land use change (Fig. 3). The combination of high radiative forcing (RCP 8.5) and high land-use change led to the steepest decline in NPP over time, driven by intense long-term drying on the leeward sides of islands under RCP 8.5 (figure S4) and sustained losses of forest and shrubland land area in the high land-use scenario (figure S5). In contrast, climate change led to increased heterotrophic respiration (R~h~) over time, such that more intense warming under RCP 8.5 (figure S4) resulted in R~h~ being ~3% higher by 2100 than under RCP 4.5 (figure 3).  Land-use change substantially reduced R~h~ in the high land-use scenario (figure 3) because of long-term reductions in forest and shrubland land area (figure S5), similar to trends in NPP. Transition-triggered carbon fluxes to the atmosphere from land use, land-use change, and wildfire were largely independent of changes in climate, stabilizing by mid-century at an average of ~0.4 Tg y^-1^ in the high land-use scenario and ~0.2 Tg y^-1^ in the low land-use scenario (figure 3). Uncertainty around transition-triggered carbon fluxes were higher in the high land-use scenario, driven primarily by greater variability in wildland fire probabilities.

@@

### *Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined over time in all four scenarios  (figure 4).  On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). In addition, the high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reduction and avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

@@

### *CO~2~ fertilization effect*
Projected estimates of both total ecosystem carbon storage and ecosystem carbon balance were highly sensitive to differing rates of a CFE on plant productivity. Under the high radiative forcing (RCP 8.5) and high land-use scenario, the inclusion of a CFE ranging from 5-15% led to ~33-98 Tg of additional carbon storage in ecosystems by the end of the century, a ~10-30% increase (figure 6a). Compared to the reference scenario (0% CFE), a 5% CFE was sufficient to transform Hawaiian Island ecosystems from a net carbon source to the atmosphere for the latter half of the 21st century (figure 4b) to a net carbon sink for the entire simulation period (figure 6b), completely offsetting all other carbon losses induced by high radiative forcing and high land use. Net carbon sink strength was further enhanced at higher CFE rates, with NBP increasing by an average of 0.07 Tg C y^-1^ for each 1% increase in CFE (figure 6b). When compared to other scenarios, applying a 5% CFE to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 0.46 ± 0.3 Tg C y^-1^, roughly equivalent to mean annual NBP in the low radiative forcing and low land-use scenario with no CFE (0.52 ± 0.12). A 15% CFE applied to the high radiative forcing and high land-use scenario resulted in a mean annual NBP  of 1.18 ± 0.29 Tg C y^-1^, more than double that of the low radiative forcing and low land-use scenario with no CFE.
